Fractional Types

### 2019-11-22 fracGC/p.tex
edit abstract, section 1-2.

\usepackage{booktabs}  
\usepackage{subcaption} 
\usepackage{amssymb}
\usepackage{stmaryrd}
\usepackage{bbm}
\usepackage{wrapfig}
\usepackage[greek,english]{babel}
\usepackage{ucs}
\usepackage[utf8x]{inputenc}
\usepackage{autofe}
\usepackage[references]{agda} 
\usepackage{newunicodechar}
\usepackage[nocenter]{qtree}
\usepackage{multicol}  
\usepackage{bbold}
\usepackage{tikz}
\usetikzlibrary{cd}
\usetikzlibrary{quotes}

@@

  We propose a novel approach based on the idea of \emph{fractional
    types}. As a simple intuitive example, allocation of a new boolean
  value initialized to \AgdaInductiveConstructor{false} also creates a value
  $\oneover{\AgdaInductiveConstructor{false}}$ that can be thought of as a garbage
  collection (gc) process specialized to reclaim, and only reclaim,
  storage containing the value $\AgdaInductiveConstructor{false}$. This gc process is a
  first-class entity that can sliced and diced by combining it with
  other gc processes and decomposing it in smaller
  processes.

@@

Instead of using the rather low-level circuit model of computation,
one can leverage the full force of type theory and category theory by
expressing reversible classical computations as \emph{isomorphisms
  over finite types}~\cite{Fiore:2004,James:2012:IE:2103656.2103667}
or \emph{equivalences over
  groupoids}~\cite{DBLP:conf/esop/CaretteS16}. This perspective is
similarly universal for reversible computation (and can be extended to
quantum computation~\cite{10.1007/978-3-319-89366-2_19}) but with the
advantage of exposing a rich mathematical structure in reversible
computations that we will exploit to solve the ``ancilla problem''
explained next.

\paragraph*{Temporary Storage using Ancilla Bits.} The universality of
the Toffoli gate for classical reversible computing and of the
universality of the Toffoli-Hadamard gates for quantum computing should
not distract from efficiency and safety concerns. The theorems proving
universality (i) assume that temporary storage (often called
\emph{ancilla bits}) may be used~\cite{Toffoli:1980}, and (ii) that
this temporary storage is returned to its initial state before
de-allocation. Indeed if no temporary storage is allowed, the Toffoli
gate is not
universal~\cite{DBLP:conf/innovations/AaronsonGS17,DBLP:journals/corr/Xu15e}
and as would be expected the more temporary storage is allowed, the
more efficient certain computations could become.

More fundamentally, the condition requiring that the temporary storage
is only de-allocated when returned to its initial condition is a
safety condition. Violating this condition destroys the symmetry
between input and output making the circuits not reversible and, in
the quantum model, causes decoherence that may destroy the quantum
state. As reviewed in Sec.~\ref{sec:examples}, ancilla bits have a
number of critical applications and yet are poorly supported in
current reversible and quantum programming languages making them a
common source of bugs.

\paragraph*{Conservation of Information and Negative Entropy.}  
According to the conventional theory of
communication~\cite{Shannon1948}, a type with $N$ values is viewed as
an abstract system that has $N$ distinguishable states where the
amount of information contained in each state is $\log{N}$. This
entropy is a measure of information which materializes itself in
memory or bandwidth requirements when storing or transmitting elements
of this type. Thus a type with 8 elements needs 3 bits of memory for
storage or 3 bits of bandwidth for communication. The logarithmic map
implies that information contained in a composite state is the sum of
the information contained in its constituents. For example, the type
$A \times B$ where $A$ has two elements and $B$ has three elements can
be thought of a composite system consisting of two independent
unrelated subsystems.  Each state of the composite system therefore
contains $\log{(2*3)} = \log{2} + \log{3}$ bits which is the sum of
the information contained in each subsystem. If we could imagine a
\emph{fractional type} $\frac{1}{A}$, this type would introduce
negative entropy. For example, a type with ``cardinality''
$\frac{1}{8}$ has entropy $\log{\frac{1}{8}} = -3$. It is natural to
interpret this negative entropy just like we interpret ``negative
money,'' as a resource (space or bandwidth) to be repaid (reclaimed)
by some other part of the system. Indeed, we will introduce such
fractional types and use them to represent ``garbage collection
processes'' that reclaim temporary storage. Just like in the case of
``negative money'' it is critical that these fractional types be
linearly used, i.e., they cannot be duplicated or erased: this
property is however automatically enforced in a reversible
computational model.

\paragraph*{Outline.} The paper solves the ancilla problem in
reversible and quantum computation using a novel concept:
\emph{fractional types}. In the next section, we introduce the problem
of ancilla, motivate its importance, and explain the limitations of
current solutions. In Sec.~\ref{sec:pi}, we review existing work by
James et al.~\cite{rc2011,rc2012,James:2012:IE:2103656.2103667} that
introduces a reversible programming language built using isomorphism
of types. Although it is conceivable that the concept of fractional
types would apply to general-purpose languages (after some
adaptation), its natural technical definition exploits symmetries
present in categorical model of type isomorphisms. We present such a
definition in Sec.~\ref{sec:dyn} which captures and illustrates the
main novelties of fractional types and their runtime representations
as first-class gc processes. This presentation partially solves a more
general and more expressive instance of the ancilla problem as
illustrated with examples. To make the arguments regarding memory
management explicit, Sec.~\ref{sec:space} presents an abstract machine
with a heap component that grows and shrinks as ancillae are allocated
and de-allocated via fractional gc processes. In Sec.~\ref{sec:dep} we
lift ``plain'' programs manipulating ancillae to the type system in
order to exploit the power of dependent types to guarantee, for the
first time, the condition necessary for safe de-allocation of
ancillae. Technically this solution requires the introduction of
pointed types on top of which are defined both a monad and co-monad of
singleton types. The tracking of ancillae value is done within the
type system via the singleton types. A number of examples exploiting
the singleton-based type system are present to demonstrate the full
expressiveness of fractional types. We furthermore demonstrate that,
from the proofs of type safety, plain (but guaranteed to be safe)
programs can be extracted and executing without errors or runtime
checks in the conventional type system. The last section concludes
with a discussion of related work and a summary of our results.

@@

\[\begin{array}{l}
                    \frac{1}{2}\ket{a}+\frac{1}{2}\ket{b}+\frac{1}{2}\ket{c}+\frac{1}{2}\ket{d}\\
\\
\Rightarrow  \textrm{phase kickback} \\
\\
-\frac{1}{2}\ket{a}+\frac{1}{2}\ket{b}+\frac{1}{2}\ket{c}+\frac{1}{2}\ket{d}\\
\\
\Rightarrow  \textrm{diffusion} \\
\\
 (\frac{1}{2}-(-\frac{1}{2}))\ket{a}+(\frac{1}{2}-\frac{1}{2})\ket{b}+(\frac{1}{2}-\frac{1}{2})\ket{c}+(\frac{1}{2}-\frac{1}{2})\ket{d})\\
\\
\Rightarrow  \textrm{simplification} \\
\\
\ket{a}
\end{array}\]

The question, which arises repeatedly in quantum computation, is how
to implement operations such as phase kickback. Essentially we want to
concurrently perform the following action on every element of the
superposition:
\[\begin{array}{l}
  \textbf{if}~\textit{current-element}=\ket{a} \\
  \textbf{then}~\textit{negate} \\
  \textbf{else}~\textit{identity}
\end{array}\]
In the quantum world, na\"\i vely attempting to test whether the
current element is equal to $\ket{a}$ by performing a measurement would
destroy the quantum superposition defeating the entire algorithm.

Here is where ancilla qubits come to the rescue. In the words of
Matuschak and Nielsen~\cite{howgrover}, the idea is the following:
\begin{quote}
  There's a rough heuristic worth noting here, which is that you can
  often convert \verb|if-then| style of thinking into quantum
  circuits. You introduce an ancilla qubit to store the outcome of
  evaluating the \verb|if| condition. And then depending on the state
  of the ancilla, you perform the appropriate state
  manipulation. Finally, when possible you reverse the initial
  computation, resetting the ancilla to its original state so you can
  subsequently ignore it.
\end{quote}

\paragraph*{Quantum Error Correction.} The main requirement for
quantum error correction is the ability to replicate the state of a
qubit onto a number of qubits~\cite[Ch.~3]{NAP25196}. This is however
impossible to do directly by the no-cloning theorem~\cite{XXX}. Again ancilla
qubits come to the rescue: because ancilla qubits start in a known
initial state, it is possible to create a simple circuit that makes
their output state match a protected qubit without knowing or
disturbing the protected qubit. Chong et al.~\cite{sigarchblog}
illustrate the idea with a simple example:
\begin{quote}
consider a 3-qubit quantum majority code in which a logical ``0'' is
encoded as ``000'' and a logical ``1'' is encoded as ``111.''  Just as with
a classical majority code, a single bit-flip error can be corrected by
restoring to the majority value.  Unlike a classical code, however, we
can not directly measure the qubits else their quantum state will be
destroyed.  Instead, we measure syndromes from each possible pair of
qubits by interacting them with an ancilla, then measure each ancilla.
Although the errors to the qubits are actually continuous, the effect
of measuring the ancilla is to discretize the errors, as well as
inform us whether an error occurred so that it can be corrected.  
\end{quote}

@@

%%%
\subsection{Ancillae Bugs in Benchmarks.}
Despite its fundamental importance, the management of ancillae is
poorly supported in current languages, simulators, and toolkits as
demonstrated in a recent analysis of bugs in quantum
benchmarks~\cite{DBLP:conf/oopsla/HuangM18}. The analysis reveals that
lack of language support in the management of ancillae leads to
several problems. For example: 
\begin{quote}
  \textbf{Bug type 6: Incorrect composition of operations using mirroring}
  Section~4.5 discussed how bugs in deallocating ancillary qubits can
  happen due to bad parameters. Here we see how bugs in deallocating
  ancillary qubits can happen due to incorrect composition of
  operations following a mirroring pattern. For example, in Table~7,
  the operations in rows 2 and 3 are respectively mirrored and undone
  in rows~6 and~5. These lines of code need careful reversal of every
  loop and every operation.
\end{quote}

>>>>

\usepackage{booktabs}  
\usepackage{subcaption} 
\usepackage{amssymb}
\usepackage{stmaryrd}
\usepackage{bbm}
\usepackage{wrapfig}
\usepackage[greek,english]{babel}
\usepackage{ucs}
\usepackage[utf8x]{inputenc}
\usepackage{autofe}
\usepackage[references]{agda} 
\usepackage{newunicodechar}
\usepackage[nocenter]{qtree}
\usepackage{multicol}  
\usepackage{bbold}
\usepackage[show]{ed}
\usepackage{tikz}
\usetikzlibrary{cd}
\usetikzlibrary{quotes}

@@

  We propose a novel approach based on the idea of \emph{fractional
    types}. As a simple intuitive example, allocation of a new boolean
  value initialized to \AgdaInductiveConstructor{false} also creates a value
  $\oneover{\AgdaInductiveConstructor{false}}$ that can be thought of as a garbage
  collection (gc) process specialized to reclaim, and only reclaim,
  storage containing the value $\AgdaInductiveConstructor{false}$. This gc process is a
  first-class entity that can be sliced and diced by combining it with
  other gc processes and decomposing it in smaller
  processes.

@@

Instead of using the low-level circuit model of computation,
one can leverage the full force of type theory and category theory by
expressing reversible classical computations as \emph{isomorphisms
  over finite types}~\cite{Fiore:2004,James:2012:IE:2103656.2103667}
or \emph{equivalences over
  groupoids}~\cite{DBLP:conf/esop/CaretteS16}. This perspective is
similarly universal for reversible computation (and can be extended to
quantum computation~\cite{10.1007/978-3-319-89366-2_19}) but with the
advantage of exposing a rich mathematical structure in reversible
computations that we will exploit to solve the ``ancilla problem''
explained next.

\paragraph*{Temporary Storage using Ancilla Bits.} The universality of
the Toffoli gate for classical reversible computing and of the
universality of the Toffoli-Hadamard gates for quantum computing should
not distract from efficiency and safety concerns. The theorems proving
universality (i) assume that temporary storage (often called
\emph{ancilla bits}) may be used~\cite{Toffoli:1980}, and (ii) that
this temporary storage is returned to its initial state before
de-allocation. Indeed if no temporary storage is allowed, the Toffoli
gate is not
universal~\cite{DBLP:conf/innovations/AaronsonGS17,DBLP:journals/corr/Xu15e}.
And, as would be expected, the more temporary storage is allowed, the
more efficient certain computations can become.

More fundamentally, the condition requiring that the temporary storage
is only de-allocated when returned to its initial condition is a
safety condition. Violating this condition destroys the symmetry
between input and output, making the circuits irreversible and, in
the quantum model, causes decoherence that may destroy the quantum
state. As reviewed in Sec.~\ref{sec:examples}, ancilla bits have a
number of critical applications, yet are poorly supported in
current reversible and quantum programming languages making them a
common source of bugs.

\paragraph*{Conservation of Information and Negative Entropy.}  
The conventional theory of
communication~\cite{Shannon1948} views a type with $N$ values as
an abstract system that has $N$ distinguishable states; the
amount of information contained in each state is $\log{N}$. This
entropy is a measure of information which materializes itself in
memory or bandwidth requirements when storing or transmitting elements
of this type. Thus a type with 8 elements needs 3 bits of memory for
storage or 3 bits of bandwidth for communication. The information
information contained in a composite state, by definition of the
logarithm, is the sum of
the information contained in its constituents. For example, the type
$A \times B$ where $A$ has two elements and $B$ has three elements can
be thought of a composite system consisting of two independent
unrelated subsystems.  Each state of the composite system therefore
contains $\log{(2*3)} = \log{2} + \log{3}$ bits, the sum of
the information contained in each subsystem. If we could imagine a
\emph{fractional type} $\frac{1}{A}$, this type would introduce
negative entropy. For example, a type with ``cardinality''
$\frac{1}{8}$ has entropy $\log{\frac{1}{8}} = -3$. It is natural to
interpret this negative entropy just like we interpret ``negative
money,'' as a resource (space or bandwidth) to be repaid (reclaimed)
by some other part of the system. Indeed, we will introduce such
fractional types and use them to represent ``garbage collection
processes'' that reclaim temporary storage. Just like in the case of
``negative money'' it is critical that these fractional types be
linearly used, i.e., they cannot be duplicated or erased: this
property is automatically enforced in a reversible
computational model.

\paragraph*{Outline.} We solve the ancilla problem in reversible and quantum
computation using a novel concept: \emph{fractional types}. In the next
section, we introduce the problem of ancilla, motivate its importance, and
explain the limitations of current solutions. In Sec.~\ref{sec:pi}, we review
existing work%
~\cite{rc2011,DBLP:conf/esop/CaretteS16,rc2012,James:2012:IE:2103656.2103667}
that introduces a reversible programming language built using type isomorphism.
Although the concept of fractional
types might apply to general-purpose languages (after some
adaptation), its natural technical definition exploits symmetries
present in the categorical model of type isomorphisms. We present such a
definition in Sec.~\ref{sec:dyn} which captures and illustrates the
main novelties of fractional types and their runtime representations
as first-class gc processes. This presentation partially solves a more
general and more expressive instance of the ancilla problem as
illustrated with examples. 
\ednote{JC: I don't understand the previous sentence. What does
``this presentation'' refer to?  The paper as a whole, or the 
details of the fractional types? There is also a tension between
``partially solves'' and ``more general and more expressive'' which
might very well be true, but makes for a very vague sentence.}
To make the arguments regarding memory
management explicit, Sec.~\ref{sec:space} presents an abstract machine
with a heap component that grows and shrinks as ancillae are allocated
and de-allocated via garbage collection. In Sec.~\ref{sec:dep} we
lift ``plain'' programs manipulating ancillae to the type system in
order to exploit the power of dependent types to guarantee, for the
first time, the condition necessary for safe de-allocation of
ancillae. Technically this solution requires the introduction of
pointed types on top of which are defined both a monad and comonad of
singleton types. The tracking of ancillae value is done within the
type system via the singleton types. A number of examples exploiting
the singleton-based type system are presented to demonstrate the full
expressiveness of fractional types. The type safety proofs demonstrate
that plain (but guaranteed safe)
programs can be extracted and executed without errors or runtime
checks in the conventional type system. The last section concludes
with a discussion of related work and a summary of our results.

@@

\[\begin{array}{l}
\frac{1}{2}\ket{a}+\frac{1}{2}\ket{b}+\frac{1}{2}\ket{c}+\frac{1}{2}\ket{d}\\ [1.5ex]
\qquad\Rightarrow  \textrm{phase kickback} \\ [1.5ex]
-\frac{1}{2}\ket{a}+\frac{1}{2}\ket{b}+\frac{1}{2}\ket{c}+\frac{1}{2}\ket{d}\\ [1.5ex]
\qquad\Rightarrow  \textrm{diffusion} \\ [1.5ex]
 (\frac{1}{2}-(-\frac{1}{2}))\ket{a}+(\frac{1}{2}-\frac{1}{2})\ket{b}+(\frac{1}{2}-\frac{1}{2})\ket{c}+(\frac{1}{2}-\frac{1}{2})\ket{d})\\ [1.5ex]
\qquad\Rightarrow  \textrm{simplification} \\ [1.5ex]
\ket{a}
\end{array}\]

The question, which arises repeatedly in quantum computation, is how
to implement operations such as phase kickback. Essentially we want to
concurrently perform the following action on every element of the
superposition:
\[\begin{array}{l}
  \textbf{if}~\textit{current-element}=\ket{a} \\
  \textbf{then}~\textit{negate} \\
  \textbf{else}~\textit{identity}
\end{array}\]
In the quantum world, na\"\i vely attempting to test whether the
current element is equal to $\ket{a}$ by performing a measurement would
destroy the quantum superposition, defeating the entire algorithm.

Ancilla bits offer a solution.  In the words of
Matuschak and Nielsen~\cite{howgrover}, the idea is as follows:
\begin{quote}
  There's a rough heuristic worth noting here, which is that you can
  often convert \verb|if-then| style of thinking into quantum
  circuits. You introduce an ancilla qubit to store the outcome of
  evaluating the \verb|if| condition. And then depending on the state
  of the ancilla, you perform the appropriate state
  manipulation. Finally, when possible you reverse the initial
  computation, resetting the ancilla to its original state so you can
  subsequently ignore it.
\end{quote}
\ednote{Would it make sense to explicitly mention CNOT here?}

\paragraph*{Quantum Error Correction.} 
Quantum error correction requires the ability to replicate the state of a
qubit onto a number of qubits~\cite[Ch.~3]{NAP25196}. This is
impossible to do directly, by the no-cloning theorem~\cite{XXX}. Again ancilla
qubits solve the problem: because they start in a known
initial state, it is possible to create a simple circuit that makes
their output state match a protected qubit without knowing or
disturbing the protected qubit. Chong et al.~\cite{sigarchblog}
illustrate the idea with a simple example:
\begin{quote}
Consider a 3-qubit quantum majority code in which a logical ``0'' is
encoded as ``000'' and a logical ``1'' is encoded as ``111.''  Just as with
a classical majority code, a single bit-flip error can be corrected by
restoring to the majority value.  Unlike a classical code, however, we
can not directly measure the qubits else their quantum state will be
destroyed.  Instead, we measure syndromes from each possible pair of
qubits by interacting them with an ancilla, then measure each ancilla.
Although the errors to the qubits are actually continuous, the effect
of measuring the ancilla is to discretize the errors, as well as
inform us whether an error occurred so that it can be corrected.  
\end{quote}

@@

%%%
\subsection{Ancillae Bugs in Benchmarks.}
Despite its fundamental importance, the management of ancillae is
poorly supported in current languages, simulators, and toolkits. 
A recent analysis of bugs in quantum
benchmarks~\cite{DBLP:conf/oopsla/HuangM18} demonstrates this. The analysis
reveals that lack of language support in the management of ancillae leads to
several problems. For example: 
\begin{quote}
  \textbf{Bug type 6: Incorrect composition of operations using mirroring}
  Section~4.5 discussed how bugs in deallocating ancillary qubits can
  happen due to bad parameters. Here we see how bugs in deallocating
  ancillary qubits can happen due to incorrect composition of
  operations following a mirroring pattern. For example, in Table~7,
  the operations in rows 2 and 3 are respectively mirrored and undone
  in rows~6 and~5. These lines of code need careful reversal of every
  loop and every operation.
\end{quote}
